---
title: "Abalone Data Analysis #2"
author: "Kramarczyk, John"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```
-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)
library(ggpubr)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1:  ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}
par(mfrow = c(1,2),  oma = c(0, 0, 2, 0))

rock_skew <- rockchalk::skewness(mydata$RATIO)
rock_kurt <- rockchalk::kurtosis(mydata$RATIO, excess = FALSE)

hist(mydata$RATIO, 
        col = "blue",
        xlab = "Ratio",
        ylab = "Frequency",
        main = "Ratio Histogram")

qqnorm(mydata$RATIO,
       main = "Ratio QQ Plot", 
       col = ifelse(mydata$RATIO %in%
c(boxplot.stats(mydata$RATIO)$out), "red", "blue"))
qqline(mydata$RATIO)

cat("Rockchalk skewness: ", rock_skew, "\n")
cat("Rockchalk kurtosis: ", rock_kurt)

main_header <- c("Rockchalk Skewness =", round(rock_skew, 4), 
                 "Rockchalk Kurtosis =", round(rock_kurt, 4))
mtext(paste(main_header, sep=" ", collapse="  "), outer = TRUE, cex = 1)
```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 13, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

rock_skew2 <- rockchalk::skewness(mydata$L_RATIO)
rock_kurt2 <- rockchalk::kurtosis(mydata$L_RATIO, excess = FALSE)
cat("Rockchalk Skewness: ", rock_skew2, "\n")
cat("Rockchalk Kurtosis: ", rock_kurt)

par(mfrow = c(2,2), oma = c(0, 0, 2, 0))

hist(mydata$L_RATIO, 
      col = "cornflowerblue",
      xlab = "L_Ratio",
      ylab = "Frequency",
      main = "L_Ratio Histogram")

qqnorm(mydata$L_RATIO,
       main = "L_Ratio QQ Plot",
       col = ifelse(mydata$L_RATIO %in%
                    c(boxplot.stats(mydata$L_RATIO)$out),
                    "coral",
                    "cornflowerblue"))
qqline(mydata$L_RATIO)

boxplot(mydata$L_RATIO ~ mydata$CLASS, main = "L_Ratio Boxplot",
ylab = "L_Ratio", xlab = "Class", beside = TRUE, col = 'cornflowerblue')
main_header <- c("Rockchalk Skewness =", round(rock_skew2, 4), 
                 "Rockchalk Kurtosis =", round(rock_kurt, 4))
mtext(paste(main_header, sep=" ", collapse="  "), outer = TRUE, cex = 1)
```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(L_RATIO ~ CLASS, data = mydata)
```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: Based on steps 1.a, 1.b and 1.c, the variable L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes. L_Ratio fails to reject the null hypothesis with the Bartlett test. There is less skew in the histogram and QQ Plot and the outliers are more evenly distributed in the QQ Plot and boxplots. Ratio in the QQ-Plot deviates from normality which is confirmed by the skewness and kurtosis whereas the kurtosis and skewness in L_Ratio are closer to a normal distribution. ***

, and failing to reject the null hypothesis with the Bartlett test of homogeneity of variances.**

***#### Section 2  ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
summary(aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata))
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))
```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The interaction term CLASS:SEX is insignificant as revealed by the p-value (0.86709). CLASS and SEX are statistically significant with L_Ratio, but the interaction between the two variables is not significant.***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(aov(L_RATIO ~ CLASS + SEX, data = mydata))
```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: Coefficients across all age classes are significant except A2-A1. The p-value of Infant and Male or Female rejects the null hypothesis that Male and Female have no significant difference. This indicates that the Male and Female groups can be combined into one category/group like Adults.***


***####  Section 3:  ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)

```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}
par(mfrow = c(1,2))

hist(mydata$VOLUME[mydata$TYPE == 'I'],
     main = "I",
     xlim = c(0,1000),
     ylim = c(0,150),
     xlab = "VOLUME",
     ylab = "count",
     col = "grey"
     )

hist(mydata$VOLUME[mydata$TYPE == 'ADULT'],
     main = "ADULT",
     xlim = c(0,1000),
     ylim = c(0,150),
     xlab = "Volume",
     ylab = "count",
     col = "grey"
     )

```

**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: Based on Volume, it does not appear that there will be any difficulties separating Infants from Adults because for the most part, the adult distribution is normally distributed and the Infant distribution is right skewed. Also, the majority of the Adult volume is between 300 – 500 whereas the majority of the Infant volume is less than 300.***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}
mydata <- mydata %>% mutate(L_SHUCK = log10(SHUCK), L_VOLUME = log10(VOLUME))
x <- ggplot(mydata, aes(VOLUME, SHUCK))
y <- ggplot(mydata, aes(L_VOLUME, L_SHUCK))

gridExtra::grid.arrange(
  x + geom_point(aes(fill = CLASS, colour=CLASS), shape = 16, alpha = 0.50)+ theme(legend.position="bottom"),
  y + geom_point(aes(fill = CLASS, colour=CLASS), shape = 16, alpha = 0.50) + theme(legend.position="bottom"),
  x + geom_point(aes(fill = TYPE, colour=TYPE), shape = 16, alpha = 0.50) + theme(legend.position="bottom"),
  y + geom_point(aes(fill = TYPE, colour=TYPE), shape = 16, alpha = 0.50) + theme(legend.position="bottom")
)
```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: Initially, there is a lot of overlap between VOLUME and SHUCK making it difficult to distinguish between CLASS or TYPE. However, after the log-transformation, it becomes easier to see the differences as both the CLASS and TYPE levels become more clustered in the top right of the plot instead of the bottom left. For example, we can now see that A1 has the lowest SHUCK weight and VOLUME, and A3 and A4 are similar in VOLUME but A3 has higher SHUCK weight. It is also important to note that although the log-transformation increased the visibility of the variability of the plot, the linear relationship did not change.***


***####   Section 4:  ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}
model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)
```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The estimated coefficients suggest a larger decrease in L_SHUCK as CLASS increases. Combining this with the earlier displays suggests that L_SHUCK increases more significantly at lower CLASSes. For example, CLASSes A2-A5 have negative coefficients, which means that with age, SHUCK weight growth slows down while VOLUME grows at a faster rate.***


**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: TYPE is not an important predictor in this regression because it does not have a strong coefficient to predict L_SHUCK. For example, in comparison to CLASS, TYPE is unlikely to predict L_SHUCK. Although TYPE is fairly significant, other independent variables are more likely to contribute to predictions of L_SHUCK for harvesting decisions, such as CLASS.***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5:  ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}
s  <- rockchalk::skewness(model$residuals)
s
k  <- rockchalk::kurtosis(model$residuals, excess = FALSE)
k
k2 <- rockchalk::kurtosis(model$residuals)
k2

par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))

hist(model$residuals,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "steelblue"
     )

qqnorm(model$residuals,
       main = "Q-Q Plot",
       col = ifelse(model$residuals %in% c(boxplot.stats(model$residuals)$out), 
                    "red", "steelblue"))

qqline(model$residuals)
```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 10, fig.height = 6}
mydata$RESIDUALS <- model$residuals

gridExtra::grid.arrange(
  
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + 
    geom_point(aes(fill = CLASS, colour=CLASS), shape = 16) + theme(legend.position="bottom"),
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) +
    geom_point(aes(fill = TYPE, colour=TYPE), shape = 16) + theme(legend.position="bottom"),
  ggplot(mydata, aes(CLASS, RESIDUALS)) +
    geom_boxplot(aes(fill = CLASS)) + theme(legend.position="none"),
  ggplot(mydata, aes(TYPE, RESIDUALS)) +
    geom_boxplot(aes(fill = TYPE)) + theme(legend.position="none")
)

bartlett.test(RESIDUALS ~ CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: For the most part, the residuals appear to be evenly distributed and close to 0. From this, VOLUME might be useful for harvesting decisions. It is important to note that the large cluster and wide distribution to the right of the scatterplots (greater than 2) indicate room for improvements to the model, harvesting procedures, etc.***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6:  ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

head(prop.infants)
head(prop.adults)
head(volume.value)
```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b, fig.width = 10, fig.height = 6}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvest ed
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants #0

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:
(med_inf_vol <- median(mydata$VOLUME[mydata$TYPE=="I"]))

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /
  total.infants 

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /
  total.adults

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants and a little more than 93% of our adults.


# Add code to determine the median adult volume:
(med_adult_vol <- median(mydata$VOLUME[mydata$TYPE=="ADULT"]))

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adult_vol) /
  total.infants

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adult_vol) /
  total.adults

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}
ggplot() +
  geom_line(aes(volume.value, prop.infants, color = "Infants")) + 
  geom_line(aes(volume.value, prop.adults, color = "Adults")) + theme(legend.position = c(0.9, 0.8), legend.title = element_blank()) + theme(legend.background=element_blank())+
  geom_vline(xintercept = med_inf_vol) +
  annotate("text",
           label = paste(round(med_inf_vol, 1), "\nMedian Infant"), 
           x = med_inf_vol + 57, y = 0.43) +
  
  geom_vline(xintercept = med_adult_vol) +
  annotate("text", label = paste(round(med_adult_vol, 1), "\nMedian Adult"), 
           x = med_adult_vol + 57, y = 0.43) + 
  
  geom_vline(xintercept = max_inf_vol) +
  annotate("text", label = paste(round(max_inf_vol, 1), "\nLargest infant"), 
           x = max_inf_vol + 57, y = 0.94) +
  geom_hline(yintercept = 0.5, linetype="dashed") +
  
  labs(x = "VOLUME", y = "PROPORTION", 
       title = "Proportion of Adults and Infants Protected")


```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The values suggest good cutoff points between the two splits (Median Infant and Median Adult). This would reduce both true negatives and false positives***

-----

More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7:   ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
harvest_diff <- ((1 - prop.adults) - (1 - prop.infants))
head(harvest_diff)
```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}

max_diff <- volume.value[which.max(smooth.difference)]
c7 <- ggplot() +
  geom_line(aes(volume.value, harvest_diff)) +
  geom_line(aes(volume.value, smooth.difference), linetype="longdash", color="darkred") +
  geom_vline(xintercept = 262.143) +
  annotate("text", label = paste('Volume =', round(max_diff, 3)), 
           x = max_diff + 20, y = 0.35, angle = 90) +
  labs(x = "Volume", y = "Difference in Proportions Harvested")

d6 <- ggplot() +
  geom_line(aes(volume.value, prop.infants, color = "Infants")) + 
  geom_line(aes(volume.value, prop.adults, color = "Adults")) + theme(legend.position = c(0.9, 0.8), legend.title = element_blank()) + theme(legend.background=element_blank())+
  geom_vline(xintercept = med_inf_vol) +
  annotate("text",
           label = paste(round(med_inf_vol, 1), "\nMed inf"), 
           x = med_inf_vol + 57, y = 0.43) +
  geom_vline(xintercept = med_inf_vol + 128.3) +
  annotate("text",
           label = paste(round(max_diff, 1), "\nMax diff"), 
           x = max_diff + 60, y = 0.8) +
  geom_vline(xintercept = med_adult_vol) +
  annotate("text", label = paste(round(med_adult_vol, 1), "\nMed adl"), 
           x = med_adult_vol + 65, y = 0.43) + 
  geom_vline(xintercept = max_inf_vol) +
  annotate("text", label = paste(round(max_inf_vol, 1), "\nLargest inf"), 
           x = max_inf_vol + 75, y = 0.94) +
  geom_hline(yintercept = 0.5, linetype="dotted") +
  labs(x = "Volume", y = "Proportion", 
       title = "Proportion of Adults and Infants Protected")

grid.arrange(c7, d6, ncol=2)
```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}
(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

(1 - prop.infants)[which.max(smooth.difference)] # 0.1764706
#OR,
sum(mydata[mydata$TYPE == "I", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.infants # [1] 0.1764706
```
-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8:   ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

a1_i <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

paste("Cutoff:", a1_i)
paste("True Positive Rate:", (1 - prop.adults)[which(volume.value == a1_i)])
paste("False Positive Rate:", (1 - prop.infants)[which(volume.value == a1_i)])
```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}

ggplot() +
  geom_line(aes(volume.value, prop.infants, color = "Infants")) + 
  geom_line(aes(volume.value, prop.adults, color = "Adults")) + theme(legend.position = c(0.9, 0.8), legend.title = element_blank()) + theme(legend.background=element_blank())+
  geom_vline(xintercept = med_inf_vol) +
  annotate("text",
           label = paste(round(med_inf_vol, 1), "\nMed inf"), 
           x = med_inf_vol + 30, y = 0.45) +
  geom_vline(xintercept = a1_i) +
  annotate("text",
           label = paste(round(a1_i, 1), "\nZero A1", "\ninf"), 
           x = a1_i + 30, y = 0.63) +
  geom_vline(xintercept = max_diff) +
  annotate("text",
           label = paste(round(max_diff, 1), "\nMax diff"), 
           x = max_diff + 30, y = 0.8) +
  geom_vline(xintercept = med_adult_vol) +
  annotate("text", label = paste(round(med_adult_vol, 1), "\nMed adl"), 
           x = med_adult_vol + 35, y = 0.45) + 
  geom_vline(xintercept = max_inf_vol) +
  annotate("text", label = paste(round(max_inf_vol, 1), "\nLargest inf"), 
           x = max_inf_vol + 38, y = 0.96) +
  geom_hline(yintercept = 0.5, linetype="dotted") +
  labs(x = "Volume", y = "Proportion", 
       title = "Proportion of Adults and Infants Protected")
```


***#### Section 9:  ####***
```{r}
max_diff <- volume.value[which.max(smooth.difference)]
a1_i <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # 206.786
```

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}
tpr_med_adults <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adult_vol) / total.adults, 3)
fpr_med_adults <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adult_vol) / total.infants, 3)

tpr_med_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) / total.adults, 3)
fpr_med_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) / total.infants, 3)

tpr_protect_all_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) / total.adults, 3)
fpr_protect_all_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) / total.infants, 3)

cutoffs <- which(volume.value %in% c(max_diff, a1_i))

ggplot(mapping = aes((1 - prop.infants), (1 - prop.adults))) + 
  
  geom_line(color = 'darkblue', size = 1) + 
  geom_abline(intercept = 0, slope = 1, color = 'darkred', linetype = 2) +
  
  geom_point(mapping = aes((1 - prop.infants[cutoffs]), (1 - prop.adults[cutoffs])), shape = 21, size = 3) +
  geom_point(mapping = aes(fpr_med_adults, tpr_med_adults), shape = 21, size = 3) +
  geom_point(mapping = aes(fpr_med_inf, tpr_med_inf), shape = 21, size = 3) +
  geom_point(mapping = aes(fpr_protect_all_inf, tpr_protect_all_inf), shape = 21, size = 3) +
  
  annotate("text", label = paste('Median infant \nvol =', round(med_inf_vol, 1)), x = 0.58, y = 0.90, size = 3) +
  annotate("text", label = paste('Zero A1 infants \nvol =', round(a1_i, 1)), x = 0.25, y = 0.90, size = 3) +
  annotate("text", label = paste('Max difference \nvol =', round(max_diff, 1)), x = 0.25, y = 0.68, size = 3) +
  annotate("text", label = paste('Median adult \nvol =', round(med_adult_vol, 1)), x = 0.12, y = 0.46, size = 3) +
  annotate("text", label = paste('Protect all infants \nvol =', round(max_inf_vol, 1)), x = 0.09, y = 0.20, size = 3) +
  
  labs(title = "ROC curve of adult and infant harvest proportions", x = "Infant harvest proportion", y = "Adult harvest proportion") + 
  theme(axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5))

```


(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
area <- round(flux::auc(x = (1 - prop.infants), y = (1 - prop.adults)), 3)
cat("Area under the curve (AUC): ", area)
```

***#### Section 10:  ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults),
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	

#Volume Rounded
round_max <- round(max_inf_vol, 3)
round_med_inf <- round(med_inf_vol, 3)
round_med_adult <- round(med_adult_vol, 3)
round_max_diff <- round(max_diff, 3)
round_a1_i <- round(a1_i, 3)

#TPR
tpr_protect_all_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) / total.adults, 3)
tpr_med_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) / total.adults, 3)
tpr_med_adults <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adult_vol) / total.adults, 3)
tpr_max_diff <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_diff) / total.adults, 3)
tpr_a1_i <- round(sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > a1_i) / total.adults, 3)

#FPR
fpr_protect_all_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) / total.infants, 3)
fpr_med_inf <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) / total.infants, 3) 
fpr_med_adults <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adult_vol) /total.infants, 3)
fpr_max_diff <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > max_diff) / total.infants, 3)
fpr_a1_i <- round(sum(mydata$VOLUME[mydata$TYPE == "I"] > a1_i) / total.infants, 3)

#totalHarvest
totalhar_protect_inf <- round(sum(mydata$VOLUME > max_inf_vol) / nrow(mydata), 3)
totalhar_med_inf <- round(sum(mydata$VOLUME > med_inf_vol) / nrow(mydata), 3)
totalhar_med_adult <- round(sum(mydata$VOLUME > med_adult_vol) / nrow(mydata), 3)
totalhar_max_diff <- round(sum(mydata$VOLUME > max_diff) / nrow(mydata), 3)
totalhar_a1_i <- round(sum(mydata$VOLUME > a1_i) / nrow(mydata), 3)

cutoff_table <- tribble(
  ~" ", ~Volume, ~TPR, ~FPR, ~totalHarvest,
  "Protect all Infants", round_max, tpr_protect_all_inf, fpr_protect_all_inf, totalhar_protect_inf,
  "Median Infants", round_med_inf, tpr_med_inf, fpr_med_inf, totalhar_med_inf,
  "Median Adults", round_med_adult, tpr_med_adults, fpr_med_adults, totalhar_med_adult,
  "Max difference", round_max_diff, tpr_max_diff, fpr_max_diff, totalhar_max_diff,
  "Zero A1 infants", round_a1_i, tpr_a1_i, fpr_a1_i, totalhar_a1_i
)
cutoff_table
```

**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: Answer: The Protect All Infants cutoff yields the lowest proportion of abalones but has the highest volume. The Median Infants cutoff has the lowest volume but has the highest proportion of abalones. The Median Adults cutoff has the lowest false positive rate. The Max Difference cutoff has a low proportion of both types. The Zero A1 Infants cutoff has the highest false positive rate. ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?   Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: 1.	I might suggest using the "Max Difference"" cutoff to help sustain the population or perhaps the "Protect All Infants"" cutoff since it has 0 FPR. But I would probably not make a specific recommendation - I would outline the risks of using the false positive measurements to describe the effects on the populations. 
2.	I would present that my analysis is limited due to the data collection methods. For example, the data was not normally distributed and there were many outliers, such as the large clustesrs and wide distribution to the right of the scatterplots (greater than 2) in 5(b). This makes it difficult to accurately assess the abalone classes, which, as a result, limits the scope of my analysis and suggests the need for further observations and studies. 
3.	If it is necessary to proceed with the current analysis, I would suggest that since young Infants are needed for sustainable abalone growth, the "Protect All Infants"" cutoff should be used to avoid overharvesting.
4.	I would suggest developing better harvesting procedures, measurement guides, etc., in order to measure as many variables as possible to further reduce false positive rates and true negatives. Additional variables could include location, and environmental and habitat factors like weather, pollution, disease, loss of habitat, predation. These were concerns developed from Data Analysis #1 and is again confirmed by this analysis.
***
